Spatial Coordination Of Stomatal Patterning Between Leaf Surfaces In Amphistomatous \emph{Arabidopsis thaliana} Incurs No Photosynthetic Advantage

### 2023-10-17 ms/ms.Rmd
working on ms results

```{r}
source("r/header.R")
fit = readr::read_rds("objects/fit.rds")
pm_pars = readr::read_csv("raw-data/2d-pm-parameters.csv", show_col_types = FALSE)
model_var = readr::read_rds("objects/model_var.rds")
```

@@

Stomatal density of *Arabidopsis thaliana* the 132 leaves measured ranged from 12 to 93 (units) with high light leaves ranging from 93 to 55 (units), medium light from 15 to 35 (units), and low light from 12 to 42 (units). Leaves were amphistomatous with a mean stomatal ratio of 0.45.

@@

## Single Surface Results

If our hypotheses that natural selection will act to reduce the diffusion path length of CO$_2$ while minimizing stomatal number are correct, then we would expect to find leaf surfaces with uniformly distributed stomata. However, to the contrary, we find that though 57 of the 132 (43.1%) leaf surfaces were significantly more uniformly dispersed than uniform random synthetic stomatal grids ($\alpha$ = 0.05), none of the leaf surfaces exhibited perfectly uniform stomatal patterning (dispersion index = 1) (\autoref{fig:single-surface}). Additionally, we hypothesized that as CO$_2$ is more limiting to photosynthesis under high light, stomata would be more uniformly dispersed in plants grown in high light than plants grown in low and medium light. The data also fail to support this hypothesis as there is no strong, discernible trend between light and stomatal patterning. Interestingly, adaxial leaf surfaces were more uniformly dispersed than associated abaxial leaf surfaces across all light treatments (F$_\text{1,126}$ = 28.8; p-value < 0.001). Rather than regulate stomatal patterning in response to light regimes, plants respond by increasing stomatal density (\autoref{fig:density}). Stomatal density drastically increased in plants grown under high light, validating the long held hypothesis that light strongly influences stomatal density (F$_\text{2,126}$ = 680.7; p-value < 0.001).

@@

### Example output

We wrapped the above functions to adjust environmental and anatomical variables. We're going to simulate over a grid of parameters, but here is what the output looks like plotted:

>>>>

```{r}
source("r/header.R")
fit = readr::read_rds("objects/fit.rds")
single_surface_results = readr::read_rds("objects/single_surface_results.rds")
pm_pars = readr::read_csv("raw-data/2d-pm-parameters.csv", show_col_types = FALSE)
model_var = readr::read_rds("objects/model_var.rds")

# Summarize results on NNI
nni = single_surface_results |>
  dplyr::transmute(sig = p_value < 0.05) |>
  dplyr::summarize(n1 = sum(sig), n2 = sum(!sig))

@@

Stomatal density of *Arabidopsis thaliana* the `r nrow(single_surface_results)` leaf surfaces measured ranged from 12 to 93 (units) with high light leaves ranging from 93 to 55 (units), medium light from 15 to 35 (units), and low light from 12 to 42 (units). Leaves were amphistomatous with a mean stomatal ratio of 0.45.

@@

<!-- Copying Q's from Intro 
1) a uniform distribution of stomata on both abaxial and adaxial leaf surfaces and 2) coordinated stomatal spacing on each surface that offsets the position of stomata (\autoref{fig:ideal-amphi-grid}). Coordination between leaf surfaces is defined in this study as the occurrence of stomata in areas farthest from stomata on the opposite leaf surface. Additionally, because CO$_2$ is more limiting for photosynthesis under high light, we hypothesize that in high light 3) there should be more stomata, and 4) stomata should be more uniformly distributed than in low light. Finally, as stomatal densities are selected for optimal operational aperture, we hypothesize that 5) stomatal length will be positively correlated with the area of the leaf surface to which it is closest. We refer to this as the 'stomatal zone', the leaf area surrounding a focal stomate closest to that stomate and therefore the zone it supplies with CO$_2$). This way, each stomate can be optimally sized relative to the mesophyll volume it supplies.
-->

## Stomatal distribution is nonrandom, but far from ideal

<!-- WORKING HERE -->
Many leaf surface (`r nni[['n1']]` of `r nni[['n1']] + nni[['n2']]`, `r round(100 * nni[['n1']] / (nni[['n1']] + nni[['n2']]), 1)`%) are significantly overdispersed compared to a random uniform distribution, but none were close to an ideal hexagonal pattern (dispersion index = 1; \autoref{fig:single-surface}). Dispersion index differs significantly among light treatments (ANOVA, $F = `r$) is significantly greater Additionally, we hypothesized that as CO$_2$ is more limiting to photosynthesis under high light, the dispersion index would closer to 1 in high light...WORKING HERE would be more uniformly dispersed in plants grown in high light than plants grown in low and medium light. The data also fail to support this hypothesis as there is no strong, discernible trend between light and stomatal patterning. Interestingly, adaxial leaf surfaces were more uniformly dispersed than associated abaxial leaf surfaces across all light treatments (F$_\text{1,126}$ = 28.8; p-value < 0.001). Rather than regulate stomatal patterning in response to light regimes, plants respond by increasing stomatal density (\autoref{fig:density}). Stomatal density drastically increased in plants grown under high light, validating the long held hypothesis that light strongly influences stomatal density (F$_\text{2,126}$ = 680.7; p-value < 0.001).

@@

### 2023-10-17 ms/ms.tex
working on ms results

\hypertarget{single-surface-results}{%
\subsection{Single Surface Results}\label{single-surface-results}}

If our hypotheses that natural selection will act to reduce the
diffusion path length of CO\(_2\) while minimizing stomatal number are
correct, then we would expect to find leaf surfaces with uniformly
distributed stomata. However, to the contrary, we find that though 57 of
the 132 (43.1\%) leaf surfaces were significantly more uniformly
dispersed than uniform random synthetic stomatal grids (\(\alpha\) =
0.05), none of the leaf surfaces exhibited perfectly uniform stomatal
patterning (dispersion index = 1) (\autoref{fig:single-surface}).
Additionally, we hypothesized that as CO\(_2\) is more limiting to
photosynthesis under high light, stomata would be more uniformly
dispersed in plants grown in high light than plants grown in low and
medium light. The data also fail to support this hypothesis as there is
no strong, discernible trend between light and stomatal patterning.
Interestingly, adaxial leaf surfaces were more uniformly dispersed than
associated abaxial leaf surfaces across all light treatments
(F\(_\text{1,126}\) = 28.8; p-value \textless{} 0.001). Rather than
regulate stomatal patterning in response to light regimes, plants
respond by increasing stomatal density (\autoref{fig:density}). Stomatal
density drastically increased in plants grown under high light,
validating the long held hypothesis that light strongly influences
stomatal density (F\(_\text{2,126}\) = 680.7; p-value \textless{}
0.001).

@@

\hypertarget{example-output}{%
\subsubsection{Example output}\label{example-output}}

We wrapped the above functions to adjust environmental and anatomical
variables. We're going to simulate over a grid of parameters, but here
is what the output looks like plotted:

>>>>

\hypertarget{stomatal-distribution-is-nonrandom-but-far-from-ideal}{%
\subsection{Stomatal distribution is nonrandom, but far from
ideal}\label{stomatal-distribution-is-nonrandom-but-far-from-ideal}}

Many leaf surface (57 of 132, 43.2\%) are significantly overdispersed
compared to a random uniform distribution, none of the leaf surfaces
exhibited perfectly uniform stomatal patterning (dispersion index = 1)
(\autoref{fig:single-surface}). Additionally, we hypothesized that as
CO\(_2\) is more limiting to photosynthesis under high light, stomata
would be more uniformly dispersed in plants grown in high light than
plants grown in low and medium light. The data also fail to support this
hypothesis as there is no strong, discernible trend between light and
stomatal patterning. Interestingly, adaxial leaf surfaces were more
uniformly dispersed than associated abaxial leaf surfaces across all
light treatments (F\(_\text{1,126}\) = 28.8; p-value \textless{} 0.001).
Rather than regulate stomatal patterning in response to light regimes,
plants respond by increasing stomatal density (\autoref{fig:density}).
Stomatal density drastically increased in plants grown under high light,
validating the long held hypothesis that light strongly influences
stomatal density (F\(_\text{2,126}\) = 680.7; p-value \textless{}
0.001).

@@
